The XAI Paradox: Systems that Perform Well for the Wrong Reasons

### 2018-08-29 Paper.tex
Details

\frenchspacing  %Required
\setlength{\pdfpagewidth}{8.5in}  %Required
\setlength{\pdfpageheight}{11in}  %Required
%PDF Info Is Required:
  \pdfinfo{
/Title (The XAI Paradox: Systems that Perform Well for the Wrong Reasons)
/Author (Cor Steging, Lambert Schomaker, Bart Verheij)}
\setcounter{secnumdepth}{0}  
 \begin{document}
% The file aaai.sty is the style file for AAAI Press 
% proceedings, working notes, and technical reports.
%
\title{The XAI Paradox: Systems that Perform Well for the Wrong Reasons}
\author{Cor Steging, Lambert Schomaker, Bart Verheij\\
Department of Artificial Intelligence, Bernoulli Institute\\
University of Groningen\\
}
\maketitle
\begin{abstract}
Successful machine learning approaches are often regarded as ``black box'' systems: they perform well, but are unable to explain the reasoning behind their decisions. The emerging sub-field of Explainable Artificial Intelligence (XAI) aims to create systems that can provide explanations underlying their internal reasoning. In this study, we investigate to what extent systems learn explanatory structures that make sense in a domain. Using artificial datasets whose internal structure is determined beforehand, this study shows that the reasoning of systems that perform well is not necessarily sound. We generate datasets whose outputs are based on predetermined formal input conditions defined in terms of symmetric Boolean functions. When the output of the generated datasets are dependent on a conjunctions of input conditions, we find that trained systems perform well on the conjunction problem, but---paradoxically---not on the individual conditions separately. This implies that a confounding
%Using artificial datasets whose internal structure is known beforehand, this study shows that the reasoning of systems that perform well is not necessarily sound. In particular, when multiple combined conditions define a dataset, systems can perform well on the combined problem and not learn each of the individual conditions. Instead, a confounding 
structure within the data is learnt that still allows making the correct decisions. In other words, the systems perform well, but for the wrong reasons, providing a puzzle for the XAI movement.
%With regards to the goal of creating explainable systems, however, unsound rationales could create irrational explanations which would be problematic for the XAI movement.
%Many of the successful modern machine learning approaches can be described as ``black box'' systems; these systems perform well, but are unable to explain the reasoning behind their decisions. The emerging sub-field of Explainable Artificial Intelligence (XAI) aims to create systems that are able to explain to their users why they made a particular decision. Using artificial datasets whose internal structure is known beforehand, this study shows that the reasoning of neural networks and decisions trees that perform well are not necessarily sound. Furthermore, when multiple conditions define a dataset, systems can perform well on the combined problem yet not learn each of the individual conditions that make up the combined problem. Instead, a confounding structure within the data is learnt that still allows making the correct decisions. With regards to the goal of creating explainable systems, however, unsound rationales could create irrational explanations which would be problematic for the XAI movement.
\end{abstract}

@@

\section{Conclusion}
\noindent In the future, machine learning algorithms that simply execute a task with a high performance will not always suffice; an explanation of their decision making will be required \cite{gunning2017explainable}. This explanation, however, is dependent on the reasoning of the systems; if the reasoning is unsound, the explanation will be unsound as well. This study has shown that machine learning algorithms do not always internalize the structure of their training data as we would expect. With regards to the emerging XAI techniques and the explainable models that they generate, such unsound rationales underlying machine learning systems can form a hindrance in creating an understandable explanation.

>>>>

\frenchspacing  %Required
\setlength{\pdfpagewidth}{8.5in}  %Required
\setlength{\pdfpageheight}{11in}  %Required
%PDF Info Is Required:
  \pdfinfo{
/Title (The XAI Paradox: Systems that Perform Well for the Wrong Reasons)
/Author (Cor Steging, Lambert Schomaker, Bart Verheij)}
\setcounter{secnumdepth}{0}  
 \begin{document}
% The file aaai.sty is the style file for AAAI Press 
% proceedings, working notes, and technical reports.
%
\title{The XAI Paradox: Systems that Perform Well for the Wrong Reasons}
\author{Cor Steging, Lambert Schomaker, Bart Verheij\\
Department of Artificial Intelligence, Bernoulli Institute\\
University of Groningen\\
}
\maketitle
\begin{abstract}
Successful machine learning approaches are often regarded as ``black box'' systems: they perform well, but are unable to explain the reasoning behind their decisions. The emerging sub-field of Explainable Artificial Intelligence (XAI) aims to create systems that can provide explanations underlying their internal reasoning. In this study, we investigate to what extent systems learn explanatory structures that make sense in a domain. Using artificial datasets whose internal structure is determined beforehand, we are able to show that the reasoning of systems that perform well is not necessarily sound. We generate datasets whose outputs are based on predetermined formal input conditions defined in terms of symmetric Boolean functions. When the output of the generated datasets are dependent on a conjunctions of input conditions, we find that trained systems perform well on the conjunction problem, but---paradoxically---not on the individual conditions separately. This implies that a confounding
%Using artificial datasets whose internal structure is known beforehand, this study shows that the reasoning of systems that perform well is not necessarily sound. In particular, when multiple combined conditions define a dataset, systems can perform well on the combined problem and not learn each of the individual conditions. Instead, a confounding 
structure within the data is learnt that still allows making the correct decisions. In other words, the systems perform well, but for the wrong reasons, providing a puzzle for the XAI movement.
%With regards to the goal of creating explainable systems, however, unsound rationales could create irrational explanations which would be problematic for the XAI movement.
%Many of the successful modern machine learning approaches can be described as ``black box'' systems; these systems perform well, but are unable to explain the reasoning behind their decisions. The emerging sub-field of Explainable Artificial Intelligence (XAI) aims to create systems that are able to explain to their users why they made a particular decision. Using artificial datasets whose internal structure is known beforehand, this study shows that the reasoning of neural networks and decisions trees that perform well are not necessarily sound. Furthermore, when multiple conditions define a dataset, systems can perform well on the combined problem yet not learn each of the individual conditions that make up the combined problem. Instead, a confounding structure within the data is learnt that still allows making the correct decisions. With regards to the goal of creating explainable systems, however, unsound rationales could create irrational explanations which would be problematic for the XAI movement.
\end{abstract}

@@

\section{Conclusion}
\noindent In the future, machine learning algorithms that simply execute a task with a high performance will not always suffice; for certain domains and certain applications, an explanation of their decision making will be required \cite{gunning2017explainable}. This explanation, however, is dependent on the reasoning of the systems; if the reasoning is unsound, the explanation will be unsound as well. This study has shown that machine learning algorithms do not always internalize the structure of their training data as we would expect. With regards to the emerging XAI techniques and the explainable models that they generate, such unsound rationales underlying machine learning systems can form a hindrance in creating an understandable explanation.
